---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gamRTMB

<!-- badges: start -->
[![R-CMD-check](https://github.com/janolefi/gamRTMB/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/janolefi/gamRTMB/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Distributional (GAMLSS-style) regression where smooth terms can enter **every**
parameter of a distribution, not just the mean.

The package is mostly glue, by design:

- **mgcv** builds the bases and penalties, and `mgcv::smooth2random()`
  reparameterises the penalized coefficients as iid Gaussian random effects —
  except where the penalty is already a sparse precision matrix, which keeps it
  and uses `RTMB::dgmrf()` instead;
- **RTMB** supplies automatic differentiation and the Laplace approximation;
- **RTMBdist** supplies the log-densities, each in its own **native**
  parameterisation — a skew normal is modelled as `xi`, `omega`, `alpha`, not
  as generic location/scale/shape.

Restricted maximum likelihood is the default, effective degrees of freedom per
smooth agree with mgcv, and smoothing parameters can be shared between terms
*and* between distributional parameters.

## Installation

``` r
# install.packages("pak")
pak::pak("janolefi/gamRTMB")
```

## Example

`MASS::mcycle` measures head acceleration in a simulated motorcycle crash. The
mean is famously hard to model — and the variance changes just as much.

```{r fit}
library(gamRTMB)
data(mcycle, package = "MASS")

fit <- gamRTMB(accel ~ list(mean = ~ s(times, bs="ad", k=30, m=4),
                            sd   = ~ s(times, k = 10)),
               data = mcycle, family = fam("norm"))
summary(fit)
```

Once every parameter varies, the useful output is covariate-dependent
quantiles rather than a mean and one standard error:

```{r quantiles, fig.height = 4.6, fig.width = 7, fig.alt = "Fitted 5th, 25th, 50th, 75th and 95th percentile curves of head acceleration against time, over the mcycle data. The curves are narrow before impact, fan out widely through it, then contract again."}
plot(fit, type = "quantile")
```

Term plots, a worm plot of randomised quantile residuals, and everything else
are in `vignette("gamRTMB")`.

## What it does

| | |
|---|---|
| Families | 85, from RTMBdist plus the standard R densities; `families()` lists them, `fam()` builds one |
| Formula | `y ~ list(mean = ~ s(x), sd = ~ s(z))`, one one-sided formula per parameter, missing ones get `~1` |
| Smooths | `s()`, `te()`, `ti()`, `t2()`, `by=`, `bs="fs"`, `bs="re"`, `bs="ad"`, shrinkage bases, and `id=` to share smoothing parameters |
| Spatial | `bs="mrf"` Markov random fields over an adjacency graph, or any precision matrix via `xt=list(penalty=)`; `bs="spde"` Matern fields on an [fmesher](https://cran.r-project.org/package=fmesher) mesh. Both kept sparse, so a few thousand regions or mesh nodes is routine |
| Criterion | `REML` by default (coefficients integrated out by the same Laplace approximation), `ML`, or `qREML` — the REML criterion optimised by extended Fellner–Schall, which fits four-parameter families the other two cannot start |
| Inference | `summary()`, `vcov()`, `edf()`, `AIC()`/`BIC()`, `predict()` with standard errors |
| Diagnostics | `residuals()` gives randomised quantile residuals; `plot(type = "worm")` |
| Also | `weights`, `offset()` inside a parameter's formula, `na.action` |

## Status

Early but checked. EDF and fitted values agree with `mgcv::gam(family =
gaulss())` to three decimals, with and without shared smoothing parameters;
`predict()` on the fitting data reproduces the in-sample linear predictors to
machine precision; and quantile standard errors match both an analytic special
case and simulation from the joint posterior. The sparse GMRF route is a
reparameterisation rather than an approximation, and is checked against the
dense one: log-likelihood, EDF, AIC, fitted values, predictions and their
standard errors all agree, on Markov random fields and on ordinary smooths
alike. The SPDE precision matches the closed form
\(\tau^2(\kappa^4 C + 2\kappa^2 G_1 + G_2)\) to 5e-16, and the fitted
range and marginal standard deviation approach their true values as the
sample grows.

Smooths that penalize one coefficient vector several times over — `te()`,
`ti()`, `bs = "ad"` — take the sparse route too, and for the same reason a
Markov random field does: `smooth2random()` needs one variance per penalized
block and has none to give. What makes that legitimate is that the null space
of \(\sum_i \lambda_i S_i\) is the intersection of the individual null
spaces and so does not move with \(\lambda\), which means one corner
constraint serves every penalty at once and leaves a positive definite
precision — so the generalised determinant \(|S_\lambda|_+\), and its
notoriously delicate stable evaluation, never has to be computed. EDF agree
with `mgcv::gam` to 1e-3 and fitted values to 3e-3 on a `te()` over
interacting and over additive data, a `ti()`, and adaptive smooths on a jump
and on `MASS::mcycle`.

Not supported: `fx = TRUE`, and smooths whose unpenalized null spaces overlap
(use a shrinkage basis, `bs = "ts"`). Not yet implemented: smooth-term
p-values and 2-D smooth plots.
