---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gamRTMB

<!-- badges: start -->
<!-- badges: end -->

Distributional (GAMLSS-style) regression where smooth terms can enter **every**
parameter of a distribution, not just the mean.

The package is mostly glue, by design:

- **mgcv** builds the bases and penalties, and `mgcv::smooth2random()`
  reparameterises the penalized coefficients as iid Gaussian random effects;
- **RTMB** supplies automatic differentiation and the Laplace approximation;
- **RTMBdist** supplies the log-densities, each in its own **native**
  parameterisation — a skew normal is modelled as `xi`, `omega`, `alpha`, not
  as generic location/scale/shape.

Restricted maximum likelihood is the default, effective degrees of freedom per
smooth agree with mgcv, and smoothing parameters can be shared between terms
*and* between distributional parameters.

Most of the package was written by Claude Code.


## Installation

``` r
# install.packages("pak")
pak::pak("janolefi/gamRTMB")
```

## Example

A Gaussian location-scale model: the mean varies smoothly with `x1`, and so
does the standard deviation.

```{r example}
library(gamRTMB)

set.seed(1)
n <- 500
d <- data.frame(x1 = runif(n), x2 = runif(n))
d$y <- rnorm(n, mean = sin(2 * pi * d$x1) + d$x2^2,
             sd = exp(-1 + 0.8 * cos(2 * pi * d$x1)))

fit <- gamRTMB(y ~ list(mean = ~ s(x1) + s(x2), sd = ~ s(x1)), data = d)
fit
```

`summary()` lays this out per distributional parameter — every parameter has
its own formula, link, coefficients and smooths:

```{r summary}
summary(fit)
```

Smooth terms are reported with their effective degrees of freedom and
smoothing parameters, but deliberately without p-values: mgcv's "approximate
significance" rests on Wood's test for a term's whole coefficient block, and a
naive Wald test in its place would mislead. `edf(fit)` returns that table on
its own.

`AIC()` and `BIC()` use the log-likelihood at the fitted coefficients with the
total effective degrees of freedom — the same convention as mgcv and GAMLSS's
GAIC — not the REML criterion, which is not a likelihood and is not comparable
across mean structures. The criterion itself stays available as
`fit$objective`.

```{r aic}
c(logLik = as.numeric(logLik(fit)), df = attr(logLik(fit), "df"), AIC = AIC(fit))
```

Predictions come back as one vector per distributional parameter. New data is
pushed through `mgcv::PredictMat()` on the stored smooths, so the basis and
knots are exactly those of the fit:

```{r predict}
grid <- data.frame(x1 = seq(0, 1, length.out = 5), x2 = 0.5)
predict(fit, newdata = grid, type = "response")
```

### Any family

`fam()` derives the parameter names, links and starting values from the
density itself, so nothing has to be hand-written per distribution.
`families()` lists what is available — everything in RTMBdist that is a
univariate regression family, plus the standard densities RTMB makes
AD-aware:

```{r families}
nrow(families())
families("gamma|^norm$|zipois")
```

Parameter names are always the density's own, which is why a Gaussian is
`mean`/`sd` rather than `mu`/`sigma`. Here a skew normal, with a smooth on
all three of its parameters at once:

```{r skewnorm}
f <- fam("skewnorm2")
f

set.seed(2)
n <- 1500
d2 <- data.frame(x1 = runif(n), x2 = runif(n), x3 = runif(n))
d2$y <- RTMBdist::rskewnorm2(n,
  mean  = sin(2 * pi * d2$x1),
  sd    = exp(-0.2 + 0.5 * cos(2 * pi * d2$x2)),
  alpha = 2.5 * sin(2 * pi * d2$x3))

fit2 <- gamRTMB(y ~ list(mean = ~ s(x1), sd = ~ s(x2), alpha = ~ s(x3)),
                family = f, data = d2)
edf(fit2)
```

Parameters that are known data rather than quantities to model — the number of
binomial trials, truncation bounds — are declared separately and are not
reachable from the formula:

```{r fixed}
fam("betabinom", fixed = list(size = "trials"))
```

### Weights, offsets and missing data

Prior weights multiply each observation's log-density contribution, exactly as
in `glm()`. An `offset()` goes inside the formula of the parameter it belongs
to — per-parameter, because `sd = ~ offset(log(s))` means something quite
different from the same term on `mean`. Missing values in any model variable
are handled by `na.action`, `na.omit` by default, and the number of dropped
rows is reported.

```{r plumbing}
set.seed(4)
n <- 400
p <- data.frame(x = runif(n), E = runif(n, 1, 5))
p$y <- rpois(n, p$E * exp(sin(2 * pi * p$x)))
p$y[1:3] <- NA

fit_p <- gamRTMB(y ~ list(lambda = ~ s(x) + offset(log(E))),
                 family = fam("pois"), data = p,
                 weights = rep(c(1, 2), n / 2))
fit_p
```

### Shared smoothing parameters

`s(..., id = )` ties smooths to one smoothing parameter, exactly as in mgcv
(including the linked bases, so the shared variance means the same thing for
each term). Unlike mgcv, the group may span distributional parameters:

```{r id}
fit3 <- gamRTMB(y ~ list(mean = ~ s(x1, id = "sh"),
                         sd   = ~ s(x1, id = "sh")),
                data = d)
fit3
edf(fit3)
```

## Diagnostics

`residuals()` gives randomised quantile (pseudo) residuals via the
probability integral transform, so a QQ plot checks the whole distributional
assumption and not just the mean. Where the response has atoms the residual is
randomised within the step, using the family's declared support to get the
left limit of the CDF — `F(y-1)` for a lattice response, `F(y) - p(y)` at an
atom of a mixed one.

```{r residuals, fig.height = 3.2, fig.width = 6.4}
set.seed(6)
n <- 1200
z <- data.frame(x = runif(n))
z$y <- ifelse(runif(n) < 0.35, 0, rpois(n, exp(1.5 + sin(2 * pi * z$x))))

wrong <- gamRTMB(y ~ list(lambda = ~ s(x)), family = fam("pois"), data = z)
right <- gamRTMB(y ~ list(lambda = ~ s(x), zeroprob = ~ 1),
                 family = fam("zipois"), data = z)

par(mfrow = c(1, 2))
set.seed(1); plot(wrong, type = "worm", main = "fitted as pois")
set.seed(1); plot(right, type = "worm", main = "fitted as zipois")
```

A worm plot is the detrended QQ plot, which makes it easier to see *where*
the distribution is wrong; `type = "qq"` gives the plain version. Because
residuals for a discrete response are randomised, `nsim` overlays several
draws so the randomisation is visible rather than hidden.

The zero-inflated data fitted as a plain Poisson bends away from the line;
the same data with a `zeroprob` parameter does not. `families()` reports which
families have a CDF, and so which support residuals.

## Term plots

`plot()` draws one panel per smooth: its contribution to that parameter's
linear predictor, with a pointwise ±2 SE band. The band comes from the joint
covariance, so it includes the uncertainty in the smoothing parameters
(mgcv's `unconditional = TRUE`).

```{r terms, fig.height = 5, fig.width = 7}
plot(fit)
```

`select` picks one term out, by index or by a pattern matched against the
`parameter: term` labels. Anything in `...` reaches the underlying `plot()`,
and `bty = "n"` is a default you can override:

```{r terms-one, fig.height = 3.4, fig.width = 5}
plot(fit, select = "sd", col = "firebrick", lwd = 2,
     main = "smooth on the log sd")
```

The plotted curves come back invisibly, so a panel can be rebuilt by hand —
here against the truth used to simulate:

```{r terms-grab, fig.keep = "none"}
cv <- plot(fit, select = 1)[[1]]   # returns the curve; also draws it
str(cv)
```

```{r terms-data, fig.height = 3.4, fig.width = 5}
par(mar = c(4, 4, 1, 1))
plot(cv$x, cv$fit, type = "l", lwd = 2, ylim = c(-2, 2), bty = "n",
     xlab = "x1", ylab = "s(x1)")
polygon(c(cv$x, rev(cv$x)), c(cv$fit + 2 * cv$se, rev(cv$fit - 2 * cv$se)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(cv$x, cv$fit, lwd = 2)
lines(cv$x, sin(2 * pi * cv$x) - mean(sin(2 * pi * d$x1)),
      col = "firebrick", lty = 3, lwd = 2)
legend("topright", c("fitted", "truth"), col = c(1, "firebrick"),
       lty = c(1, 3), bty = "n")
```

`predict(se.fit = TRUE)` gives the same standard errors numerically, per term
or for the whole linear predictor, if you would rather build everything
yourself.

## Status and scope

Early but checked. EDF and fitted values agree with `mgcv::gam(family =
gaulss())` to three decimals, with and without shared smoothing parameters,
and `predict()` on the fitting data reproduces the in-sample linear predictors
to machine precision.

Supported: `s()`, `t2()`, `by =` variables, `bs = "fs"`, `bs = "re"`, and
`id =`. Not supported: `te()` (mgcv itself declines
`smooth2random(type = 2)` for it — use `t2()`) and `fx = TRUE` (no penalized
part to make random).

An extended Fellner–Schall fitting engine is planned as an alternative to the
Laplace one; `dev/NOTES-fellner-schall.md` records the derivation and the
seams it needs.
