---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gamRTMB

<!-- badges: start -->
[![R-CMD-check](https://github.com/janolefi/gamRTMB/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/janolefi/gamRTMB/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Distributional (GAMLSS-style) regression where smooth terms can enter **every**
parameter of a distribution, not just the mean.

The package is mostly glue, by design:

- **mgcv** builds the bases and penalties, and `mgcv::smooth2random()`
  reparameterises the penalized coefficients as iid Gaussian random effects;
- **RTMB** supplies automatic differentiation and the Laplace approximation;
- **RTMBdist** supplies the log-densities, each in its own **native**
  parameterisation — a skew normal is modelled as `xi`, `omega`, `alpha`, not
  as generic location/scale/shape.

Restricted maximum likelihood is the default, effective degrees of freedom per
smooth agree with mgcv, and smoothing parameters can be shared between terms
*and* between distributional parameters.

## Installation

``` r
# install.packages("pak")
pak::pak("janolefi/gamRTMB")
```

## Example

`MASS::mcycle` measures head acceleration in a simulated motorcycle crash. The
mean is famously hard to model — and the variance changes just as much.

```{r fit}
library(gamRTMB)
data(mcycle, package = "MASS")

fit <- gamRTMB(accel ~ list(mean = ~ s(times, k = 20),
                            sd   = ~ s(times, k = 12)),
               data = mcycle)
summary(fit)
```

Once every parameter varies, the useful output is covariate-dependent
quantiles rather than a mean and one standard error:

```{r quantiles, fig.height = 4.6, fig.width = 7, fig.alt = "Fitted 5th, 25th, 50th, 75th and 95th percentile curves of head acceleration against time, over the mcycle data. The curves are narrow before impact, fan out widely through it, then contract again."}
plot(fit, type = "quantile")
```

Term plots, a worm plot of randomised quantile residuals, and everything else
are in `vignette("gamRTMB")`.

## What it does

| | |
|---|---|
| Families | 85, from RTMBdist plus the standard R densities; `families()` lists them, `fam()` builds one |
| Formula | `y ~ list(mean = ~ s(x), sd = ~ s(z))`, one one-sided formula per parameter, missing ones get `~1` |
| Smooths | `s()`, `t2()`, `by=`, `bs="fs"`, `bs="re"`, shrinkage bases, and `id=` to share smoothing parameters |
| Criterion | REML by default (coefficients integrated out by the same Laplace approximation), or ML |
| Inference | `summary()`, `vcov()`, `edf()`, `AIC()`/`BIC()`, `predict()` with standard errors |
| Diagnostics | `residuals()` gives randomised quantile residuals; `plot(type = "worm")` |
| Also | `weights`, `offset()` inside a parameter's formula, `na.action` |

## Status

Early but checked. EDF and fitted values agree with `mgcv::gam(family =
gaulss())` to three decimals, with and without shared smoothing parameters;
`predict()` on the fitting data reproduces the in-sample linear predictors to
machine precision; and quantile standard errors match both an analytic special
case and simulation from the joint posterior.

Not supported: `te()` (mgcv itself declines `smooth2random(type = 2)` for it —
use `t2()`), `fx = TRUE`, and smooths whose unpenalized null spaces overlap
(use a shrinkage basis, `bs = "ts"`). Not yet implemented: smooth-term
p-values, 2-D smooth plots, and an extended Fellner–Schall fitting engine as
an alternative to the Laplace one — `dev/NOTES-fellner-schall.md` records the
derivation and the seams it needs.
