---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gamRTMB

<!-- badges: start -->
<!-- badges: end -->

Distributional (GAMLSS-style) regression where smooth terms can enter **every**
parameter of a distribution, not just the mean.

The package is mostly glue, by design:

- **mgcv** builds the bases and penalties, and `mgcv::smooth2random()`
  reparameterises the penalized coefficients as iid Gaussian random effects;
- **RTMB** supplies automatic differentiation and the Laplace approximation;
- **RTMBdist** supplies the log-densities, each in its own **native**
  parameterisation — a skew normal is modelled as `xi`, `omega`, `alpha`, not
  as generic location/scale/shape.

Restricted maximum likelihood is the default, effective degrees of freedom per
smooth agree with mgcv, and smoothing parameters can be shared between terms
*and* between distributional parameters.

## Installation

``` r
# install.packages("pak")
pak::pak("janolefi/gamRTMB")
```

## Example

A Gaussian location-scale model: the mean varies smoothly with `x1`, and so
does the standard deviation.

```{r example}
library(gamRTMB)

set.seed(1)
n <- 500
d <- data.frame(x1 = runif(n), x2 = runif(n))
d$y <- rnorm(n,
             mean = sin(2 * pi * d$x1) + d$x2^2,
             sd   = exp(-1 + 0.8 * cos(2 * pi * d$x1)))

fit <- gamRTMB(y ~ list(mean = ~ s(x1) + s(x2),
                        sd   = ~ s(x1)),
               data = d)
fit
```

Effective degrees of freedom per smooth, as in a `gam` summary:

```{r edf}
edf(fit)
```

Predictions come back as one vector per distributional parameter. New data is
pushed through `mgcv::PredictMat()` on the stored smooths, so the basis and
knots are exactly those of the fit:

```{r predict}
grid <- data.frame(x1 = seq(0, 1, length.out = 5), x2 = 0.5)
predict(fit, newdata = grid, type = "response")
```

### Any family

`fam()` derives the parameter names, links and starting values from the
density itself, so nothing has to be hand-written per distribution.
`families()` lists what is available — everything in RTMBdist that is a
univariate regression family, plus the standard densities RTMB makes
AD-aware:

```{r families}
nrow(families())
families("gamma|^norm$|zipois")
```

Parameter names are always the density's own, which is why a Gaussian is
`mean`/`sd` rather than `mu`/`sigma`. Here a skew normal, with a smooth on
all three of its parameters at once:

```{r skewnorm}
f <- fam("skewnorm2")
f

set.seed(2)
n <- 1500
d2 <- data.frame(x1 = runif(n), x2 = runif(n), x3 = runif(n))
d2$y <- RTMBdist::rskewnorm2(n,
  mean  = sin(2 * pi * d2$x1),
  sd    = exp(-0.2 + 0.5 * cos(2 * pi * d2$x2)),
  alpha = 2.5 * sin(2 * pi * d2$x3))

fit2 <- gamRTMB(y ~ list(mean = ~ s(x1), sd = ~ s(x2), alpha = ~ s(x3)),
                family = f, data = d2)
edf(fit2)
```

Parameters that are known data rather than quantities to model — the number of
binomial trials, truncation bounds — are declared separately and are not
reachable from the formula:

```{r fixed}
fam("betabinom", fixed = list(size = "trials"))
```

### Shared smoothing parameters

`s(..., id = )` ties smooths to one smoothing parameter, exactly as in mgcv
(including the linked bases, so the shared variance means the same thing for
each term). Unlike mgcv, the group may span distributional parameters:

```{r id}
fit3 <- gamRTMB(y ~ list(mean = ~ s(x1, id = "sh"),
                         sd   = ~ s(x1, id = "sh")),
                data = d)
fit3
edf(fit3)
```

## Confidence bands

Ask for the joint precision matrix at fit time and `predict()` will return
standard errors, per term or for the whole linear predictor:

```{r bands, fig.height = 3.2, fig.width = 5}
fitb <- gamRTMB(y ~ list(mean = ~ s(x1) + s(x2), sd = ~ s(x1)), data = d)

g <- data.frame(x1 = seq(0, 1, length.out = 200), x2 = 0.5)
tm <- predict(fitb, newdata = g, type = "terms", se.fit = TRUE)

par(mar = c(4, 4, 1, 1))
plot(g$x1, tm$mean$fit[, 1], type = "l", lwd = 2, ylim = c(-2, 2),
     xlab = "x1", ylab = "s(x1)")
polygon(c(g$x1, rev(g$x1)),
        c(tm$mean$fit[, 1] + 2 * tm$mean$se[, 1],
          rev(tm$mean$fit[, 1] - 2 * tm$mean$se[, 1])),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(g$x1, sin(2 * pi * g$x1) - mean(sin(2 * pi * d$x1)),
      col = 2, lty = 2, lwd = 2)
legend("topright", c("fitted", "truth"), col = c(1, 2), lty = c(1, 2), bty = "n")
```

## Status and scope

Early but checked. EDF and fitted values agree with `mgcv::gam(family =
gaulss())` to three decimals, with and without shared smoothing parameters,
and `predict()` on the fitting data reproduces the in-sample linear predictors
to machine precision.

Supported: `s()`, `t2()`, `by =` variables, `bs = "fs"`, `bs = "re"`, and
`id =`. Not supported: `te()` (mgcv itself declines
`smooth2random(type = 2)` for it — use `t2()`) and `fx = TRUE` (no penalized
part to make random).

An extended Fellner–Schall fitting engine is planned as an alternative to the
Laplace one; `dev/NOTES-fellner-schall.md` records the derivation and the
seams it needs.
