---
title: "gamRTMB and GAMLSS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gamRTMB and GAMLSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
ok <- requireNamespace("gamlss", quietly = TRUE) &&
      requireNamespace("gamlss.data", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = ok,
                      fig.width = 7, fig.height = 3.6, dpi = 110)
```

```{r skip, echo = FALSE, eval = !ok, results = "asis"}
cat("This vignette needs **gamlss** and **gamlss.data**, which are not installed.")
```

**gamlss** is the reference implementation of distributional regression, so the
first question about `gamRTMB` is whether it agrees with it. This vignette fits
three standard GAMLSS examples both ways.

```{r libs, message = FALSE}
library(gamRTMB)
library(gamlss)
library(gamlss.data)
tm <- function(e) unname(system.time(e)["elapsed"])
```

A note before the numbers: `gamlss` masks `edf()`, which `gamRTMB` also
exports, so with both attached you need `gamRTMB::edf()`.

```{r edf}
edf <- gamRTMB::edf
```

The two packages are **not** doing quite the same arithmetic. `gamlss` uses
P-splines (`pb()`) with a local maximum-likelihood smoothing criterion;
`gamRTMB` uses mgcv's bases with REML via a Laplace approximation. So the
fitted curves should agree closely, and the effective degrees of freedom
roughly, but nothing should match to the last digit.

## Normal location-scale: `abdom`

Foetal abdominal circumference against gestational age — the standard
introductory GAMLSS example.

```{r abdom}
data(abdom)

t_gl <- tm(m1 <- gamlss(y ~ pb(x), sigma.fo = ~ pb(x), data = abdom,
                        trace = FALSE))
t_rt <- tm(f1 <- gamRTMB(y ~ list(mean = ~ s(x), sd = ~ s(x)), data = abdom))

data.frame(
  fit      = c("gamlss", "gamRTMB"),
  seconds  = round(c(t_gl, t_rt), 2),
  edf      = round(c(m1$df.fit, attr(edf(f1), "edf.total")), 2),
  `-2logL` = round(c(m1$G.deviance, -2 * as.numeric(logLik(f1))), 1),
  AIC      = round(c(AIC(m1), AIC(f1)), 1),
  check.names = FALSE
)
```

Nearly identical: 7.68 against 7.63 effective degrees of freedom, and AICs
0.15 apart. The fitted curves agree to within a fraction of a percent of the
response's own spread.

```{r abdom-curves}
p1 <- predict(f1, type = "response")
c(mean = max(abs(fitted(m1, "mu")    - p1$mean)) / sd(abdom$y),
  sd   = max(abs(fitted(m1, "sigma") - p1$sd))   / sd(abdom$y))
```

## Gamma: `rent`

Munich rents against floor space. This one exposes a difference in
convention rather than in fit: `gamlss`'s `GA` family parameterises the second
argument as a **coefficient of variation**, so its `sigma` must be multiplied
by `mu` to be compared with the standard deviation `RTMBdist`'s `gamma2`
reports. `gamRTMB` always uses the density's own parameterisation, which is
the point of not renaming parameters to `mu`/`sigma`.

```{r rent}
data(rent)

t_gl <- tm(m2 <- gamlss(R ~ pb(Fl), sigma.fo = ~ pb(Fl), family = GA,
                        data = rent, trace = FALSE))
t_rt <- tm(f2 <- gamRTMB(R ~ list(mean = ~ s(Fl), sd = ~ s(Fl)),
                         family = fam("gamma2"), data = rent))

data.frame(
  fit     = c("gamlss", "gamRTMB"),
  seconds = round(c(t_gl, t_rt), 2),
  edf     = round(c(m2$df.fit, attr(edf(f2), "edf.total")), 2),
  AIC     = round(c(AIC(m2), AIC(f2)), 1)
)

p2 <- predict(f2, type = "response")
c(mean = cor(fitted(m2, "mu"), p2$mean),
  sd   = cor(fitted(m2, "sigma") * fitted(m2, "mu"), p2$sd))
```

## Four parameters: Box-Cox *t* on `abdom`

The LMS-style centile fit, and GAMLSS's flagship. `RTMBdist` implements the
same `BCT` density, to the digit:

```{r bct-dens}
c(RTMBdist = RTMBdist::dbct(60, 100, 0.1, 1, 5, log = TRUE),
  gamlss   = gamlss.dist::dBCT(60, 100, 0.1, 1, 5, log = TRUE))
```

```{r bct}
t_gl <- tm(m3 <- gamlss(y ~ pb(x), sigma.fo = ~ pb(x), nu.fo = ~ 1,
                        tau.fo = ~ 1, family = BCT, data = abdom,
                        trace = FALSE))
t_rt <- tm(f3 <- gamRTMB(y ~ list(mu = ~ s(x), sigma = ~ s(x),
                                  nu = ~ 1, tau = ~ 1),
                         family = fam("bct"), data = abdom))

data.frame(
  fit     = c("gamlss", "gamRTMB"),
  seconds = round(c(t_gl, t_rt), 2),
  edf     = round(c(m3$df.fit, attr(edf(f3), "edf.total")), 2),
  AIC     = round(c(AIC(m3), AIC(f3)), 1)
)
```

Here the two part company a little. The fitted curves still agree —

```{r bct-curves}
p3 <- predict(f3, type = "response")
c(mu = cor(fitted(m3, "mu"), p3$mu), sigma = cor(fitted(m3, "sigma"), p3$sigma))
```

— but `gamRTMB` spends about three more effective degrees of freedom and pays
for them in AIC. REML and GAMLSS's criterion are choosing different amounts of
smoothing on a 610-point dataset, and on this example GAMLSS chooses better.
Worth knowing rather than glossing over.

```{r bct-centiles, fig.height = 4.2, fig.alt = "Fitted centile curves of abdominal circumference against gestational age, from the Box-Cox t model, over the data."}
plot(f3, type = "quantile", prob = c(0.03, 0.1, 0.5, 0.9, 0.97))
```

## Speed

One run each on one machine, so read these as orders of magnitude rather than
benchmarks. Dutch boys' BMI against age, normal location-scale, at three
sample sizes:

```{r speed}
data(dbbmi)
sizes <- c(500, 2000, nrow(dbbmi))
res <- t(vapply(sizes, function(n) {
  d <- if (n == nrow(dbbmi)) dbbmi else
    { set.seed(1); dbbmi[sort(sample(nrow(dbbmi), n)), ] }
  c(gamlss  = tm(gamlss(bmi ~ pb(age), sigma.fo = ~ pb(age), data = d,
                        trace = FALSE)),
    gamRTMB = tm(gamRTMB(bmi ~ list(mean = ~ s(age), sd = ~ s(age)), data = d)))
}, c(gamlss = 0, gamRTMB = 0)))
data.frame(n = sizes, round(res, 2))
```

Neither dominates. Both stay within a factor of two of each other across this
range, and which one is ahead is not even monotone in `n` — which is itself
the finding: the differences are small enough that a single run cannot
separate them, and no scaling story should be read into three points this
close. Speed is not a reason to pick either package here.

## What actually differs

Where `gamRTMB` gives you something else:

- **Parameter names and count.** GAMLSS is capped at four parameters called
  `mu`, `sigma`, `nu`, `tau`; `gamRTMB` has no cap and uses each density's own
  names — hence `mean`/`sd` for a Gaussian and `xi`/`omega`/`alpha` for a skew
  normal. The `rent` example shows why that matters: `GA`'s `sigma` is a
  coefficient of variation, and a generic name hides it.
- **mgcv's smooths.** `s()`, `t2()`, `by=` factors, `bs="fs"`, `bs="re"`,
  shrinkage bases, and `id=` to share a smoothing parameter between terms *and
  between distributional parameters*.
- **Exact derivatives**, by automatic differentiation rather than analytic
  or numerical ones supplied per family.

Where GAMLSS is ahead today: term selection (`stepGAIC`), distribution search
(`fitDist`, `chooseDist`), censoring and truncation of arbitrary families
(`gamlss.cens`, `gamlss.tr`), finite mixtures, and twenty years of use. If you
need those, use GAMLSS.
