---
title: "Introduction to gamRTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gamRTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center", dpi = 110)
```

A regression model usually lets the mean depend on covariates and treats
everything else about the distribution as constant. That is often wrong, and
when it is, the mean is not the part that misleads you: the intervals are.

`gamRTMB` fits models where **every** parameter of a distribution can carry
smooth terms. It is mostly glue — `mgcv` builds the bases and penalties,
`RTMB` supplies automatic differentiation and the Laplace approximation, and
`RTMBdist` supplies the log-densities — so the interfaces should feel familiar
if you have used either.

```{r}
library(gamRTMB)
```

## The data

`MASS::mcycle` records head acceleration in a simulated motorcycle crash: 133
measurements of `accel` against `times`, in milliseconds after impact. It is
the standard example of a mean that is hard to model, and it also has an
obvious second feature.

```{r data, fig.alt = "Scatterplot of head acceleration against time for the mcycle data. Points lie flat and tightly clustered before about 15 ms, then swing sharply negative and scatter widely, before settling."}
data(mcycle, package = "MASS")
plot(accel ~ times, data = mcycle, pch = 20, col = "grey30", bty = "n",
     xlab = "time (ms)", ylab = "acceleration (g)")
```

Before about 15 ms nothing happens and the points sit tightly on a flat line.
Through the impact they swing violently and scatter widely, then settle. So
the **variance changes with time as clearly as the mean does**. A model that
assumes constant variance will report intervals that are far too wide early on
and too narrow at the impact.

## A first fit

The formula gives the response on the left and a `list()` of one-sided
formulas on the right, one per distributional parameter. Parameters are named
as the density names them, so a Gaussian is `mean` and `sd` rather than
generic labels:

```{r fam}
fam("norm")
```

Let the mean and the standard deviation each have their own smooth:

```{r fit1}
fit <- gamRTMB(accel ~ list(mean = ~ s(times, k = 20),
                            sd   = ~ s(times, k = 12)),
               data = mcycle)
fit
```

```{r summary1}
summary(fit)
```

Both smooths are strongly non-linear: 14.4 and 7.6 effective degrees of
freedom out of the 19 and 11 the bases allow. The smoothing parameters and
EDFs come from a REML fit and agree with `mgcv` to three decimals on models
both packages can express.

```{r terms1, fig.height = 3.4, fig.alt = "Two term plots with pointwise confidence bands: the smooth on the mean traces the crash waveform, and the smooth on the log standard deviation rises steeply through the impact."}
par(mfrow = c(1, 2))
plot(fit)
```

The right-hand panel is the point. The smooth on `sd` spans about 3.7 units on
the log scale, so the fitted standard deviation varies by a factor of roughly
40 between the quiet stretch before impact and the violent part.

## Was constant variance really inadequate?

Fit the same mean structure with a constant `sd` and compare. `AIC()` uses the
log-likelihood at the fitted coefficients with the total effective degrees of
freedom, the same convention `mgcv` and GAMLSS's GAIC use, so the two are
comparable:

```{r fit0}
fit0 <- gamRTMB(accel ~ list(mean = ~ s(times, k = 20), sd = ~ 1),
                data = mcycle)
c(constant_sd = AIC(fit0), smooth_sd = AIC(fit))
```

A difference of that size is not a close call.

## Checking the distribution, not just the mean

Randomised quantile residuals turn the whole distributional assumption into
something you can look at: if the model is right they are standard normal. The
worm plot is the detrended QQ plot, which makes it easier to see *where* a
distribution is wrong.

```{r resid, fig.height = 3.4, fig.alt = "Two worm plots. With a constant standard deviation a few lower-tail points stray outside the pointwise band; with a smooth standard deviation all points stay inside it."}
par(mfrow = c(1, 2))
set.seed(1)
plot(fit0, type = "worm", main = "constant sd")
plot(fit,  type = "worm", main = "smooth sd")
```

The difference is real but not dramatic: with `sd` held constant a few points
in the lower tail leave the band, reaching about 1.4 times its width, while
with a smooth `sd` nothing leaves it at all. Worth noting how much weaker this
evidence is than the AIC gap of 111 — with 133 observations a worm plot is a
blunt instrument, and it is the more useful of the two for telling you *where*
a fit is wrong rather than *whether*.

## Is a Gaussian the right shape at all?

With `sd` free the residuals look fine, but it is worth asking whether the
*shape* needs changing too. Switching family is one argument: `fam()` derives
the parameter names, links and starting values from any suitable density, so
nothing has to be written per distribution.

A power exponential has a third parameter `nu` controlling the tails, with
`nu = 2` being exactly Gaussian:

```{r pe}
fam("powerexp2")

fit_pe <- gamRTMB(accel ~ list(mu = ~ s(times, k = 20),
                               sigma = ~ s(times, k = 12),
                               nu = ~ 1),
                  family = fam("powerexp2"), data = mcycle)
c(nu = exp(coef(fit_pe)$beta[["nu:(Intercept)"]]))
c(gaussian = AIC(fit), power_exponential = AIC(fit_pe))
```

`nu` comes out near 3.7 rather than 2, which points to slightly *lighter*
tails than Gaussian, but AIC does not prefer the richer model. So there is no
case for changing the shape once the variance is free to vary. That is a
useful negative result, and the interesting part is the order in which you
find it: fitted with constant variance, the residuals of this data look
heavy-tailed and would tempt you towards a *t* distribution. The tails were
never the problem — the variance was.

Not every family will fit this data. Several heavier-tailed ones fail to
converge here, and `gamRTMB` says so plainly rather than returning a fit that
looks fine:

```{r nonconv, warning = FALSE}
fit_t <- gamRTMB(accel ~ list(mu = ~ s(times, k = 20),
                              sigma = ~ s(times, k = 12), df = ~ 1),
                 family = fam("t2"), data = mcycle)
fit_t$convergence
```

## What the model is for

Once every parameter varies, the useful output is not a mean and a single
standard error but covariate-dependent quantiles:

```{r quantiles, fig.height = 4.4, fig.alt = "Fitted percentile curves of acceleration against time over the data, with a shaded band on the median curve. The curves contract and fan out as the fitted spread changes."}
plot(fit, type = "quantile", prob = c(0.05, 0.25, 0.5, 0.75, 0.95))
```

The curves contract and fan out with the fitted spread, which is exactly what
the second smooth bought. The shaded band is the median curve's own estimation
uncertainty — not a prediction interval; the outer quantiles are that.

The fan summarises the fitted distribution by its quantiles. To see the shape
itself, `type = "density"` draws the fitted density turned on its side at a
few times. This works for every family, including the many that have no
quantile function:

```{r density, fig.height = 4.4, fig.alt = "Fitted conditional densities of acceleration, rotated and drawn at five times over the data. The density is a narrow spike before impact and much wider through it."}
plot(fit, type = "density")
```

`predict()` gives the quantiles as numbers, with standard errors from the
delta method across all the linear predictors at once:

```{r predq}
nd <- data.frame(times = c(10, 20, 30, 40))
q <- predict(fit, newdata = nd, type = "quantile", prob = c(0.1, 0.5, 0.9),
             se.fit = TRUE)
round(cbind(times = nd$times, q$fit), 1)
round(q$se.fit, 2)
```

Note how the standard errors themselves vary with time: the fit is far more
certain about the flat stretch than about the impact.

## Two things worth knowing

**Sharing smoothness.** `s(..., id = )` ties smooths to one smoothing
parameter, as in `mgcv`. Unlike `mgcv`, the group may span distributional
parameters — a way of saying "the mean and the spread vary on the same
scale". On this data that turns out to be a bad assumption, which is worth
seeing:

```{r id}
fit_id <- gamRTMB(accel ~ list(mean = ~ s(times, k = 20, id = "t"),
                               sd   = ~ s(times, k = 12, id = "t")),
                  data = mcycle)
edf(fit_id)
c(free = AIC(fit), tied = AIC(fit_id))
```

Forced to a common smoothness, the mean drops from 14.4 to under 3 effective
degrees of freedom while the `sd` smooth takes up 10, and AIC gets much worse.
The two functions genuinely differ in roughness. (Note also that the `k`
column changes: sharing an `id` links the *bases* as well, so both terms
inherit the first one's basis dimension.)

**Term selection.** A smooth's unpenalized null space means an ordinary `s()`
can shrink at most to a straight line. A shrinkage basis (`bs = "ts"` or
`bs = "cs"`) penalizes the null space too, so a term can shrink away
completely — useful when you are not sure a parameter needs a covariate at
all:

```{r select}
set.seed(1)
mcycle$noise <- runif(nrow(mcycle))
fit_sel <- gamRTMB(accel ~ list(mean = ~ s(times, bs = "ts", k = 20),
                                sd   = ~ s(times, bs = "ts", k = 12) +
                                        s(noise, bs = "ts", k = 8)),
                   data = mcycle)
edf(fit_sel)
```

The irrelevant covariate is driven to an effective degrees of freedom of
about 3e-07 — gone, not merely small — while the real terms are untouched. An
ordinary `s()` cannot do this: its unpenalized null space means it can shrink
at most to a straight line.

## Where to go next

`families()` lists every distribution available, with its parameters, links
and whether residuals and quantiles are supported:

```{r families}
nrow(families())
families("^norm$|^t2$|gamma2|zipois")
```

`?gamRTMB` documents the fitting options, including `weights`, `offset()`
terms inside a parameter's formula, `na.action`, and `method = "ML"` as an
alternative to the default REML.
